Hypertension and microbiome

### 2019-10-11 articletwo.Rmd
Deseq analyses

## DeSeq

```{r deseq results}
dset.deseq.signf <- results(dds$HYPERTENSION, contrast = c("HYPERTENSION", 0, 1)) %>%
    as.data.frame %>%
    tibble::rownames_to_column("Feature") %>%
    dplyr::mutate(qval = p.adjust(pvalue, method="BH")) %>%
    dplyr::filter(qval < 0.05)

dset.deseq.boxplot <- pseq.genus %>%
    microbiome::transform(., transform = "compositional") %>%
    prune_taxa(pull(dset.deseq.signf, Feature), .) %>%
    psmelt %>%
    mutate(Abundance = sqrt(Abundance))
```

```{r deseq results numerical}
dset.deseq.signf %>% kable
```

```{r deseq plot}
g.deseq <- dset.deseq.signf %>%
    ggplot(aes(x=gsub("g_", "", Feature), y=log2FoldChange, color=qval)) +
    geom_point(size=1) +
    geom_hline(yintercept = 0, color = "gray") +
    theme(axis.text.x = element_text(angle = -90, hjust = 0, vjust=0.5)) +
    coord_flip() +
    xlab("Feature") +
    theme_classic()
ggsave(file = "cache/deseq.png", plot=g.deseq, height=4, width=5)
```

```{r deseq boxplot}
g.boxplot <- dset.deseq.boxplot %>%
    group_by(OTU) %>%
    mutate(outlier = abs(Abundance) > median(Abundance) + IQR(Abundance) * 1.5) %>%
    ggplot(aes(x = HYPERTENSION, y = Abundance, color = OTU)) +
    geom_boxplot(lwd=0.1, outlier.shape = NA) +
    geom_point(data = function(x) dplyr::filter_(x, ~ outlier),
               position = position_jitterdodge(jitter.width = 0.02),
               size = 0.1,
               alpha = 0.4, shape = ".") +
    scale_y_continuous(limits = c(0, 0.2), expand = c(0, 0)) +
    scale_x_discrete(labels=c("0" = "Normotensive", "1" = "Hypertensive")) +
    ylab("Abundance (Hellinger)") +
    xlab("") +
    theme_classic() +
    guides(shape = guide_legend(override.aes = list(size = 0.1)),
           color = guide_legend(override.aes = list(size = 0.1))) +
    theme(legend.title = element_text(size = 4), 
          legend.text  = element_text(size = 4),
          legend.key.size = unit(1, "lines"),
          legend.position="bottom")
ggsave(file = "cache/dds-test.png", plot=g.boxplot, height=4, width=6, units = "in", dpi = 300)
```

@@

For magnitude, age, sex, and BMI explain `r round(summary(lm(MAP ~ BL_AGE +
SEX + BMI, data = meta(pseq.genus)))$adj.r.squared * 100, 1)`% of MAP.

```{r maaslin clean results}
maas.results.df.signf <- maaslin %>%
    merge(names.dset %>% select(Covariate, Category, Name), by.x ="Variable", by.y = "Covariate") %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("[gs]_", "", Feature),
                  beta = Coefficient) %>%
    dplyr::filter(Qval < 0.05) %>%
    arrange(Feature)
```

```{r maaslin insignif, include = FALSE, eval = FALSE}
 maaslin %>%
    dplyr::mutate(Qval = p.adjust(P.value, method="BH"),
                  Feature = gsub("[gs]_", "", Feature),
                  beta = Coefficient) %>%
     arrange(Feature) %>%
     select(Variable, Feature, Qval) %>%
     spread(Variable, Qval)
```

```{r maaslin plot}
my.plot.order <- function(x) {
    y <- seq(length(x))
    y[x == "Systolic BP"] = 4
    y[x == "Diastolic BP"] = 3
    y[x == "Pulse pressure"] = 2
    y[x == "Mean arterial pressure"] = 1
    y[x == "Hypertension"] = 0
    y
}

maas.all.pairs <- expand.grid(Feature = maas.results.df.signf %>% pull(Feature),
                         Name = maas.results.df.signf %>% pull(Name))

maas.tables.df <- merge(maas.results.df.signf,
                        maas.all.pairs,
                        all.y=TRUE,
                        by.y=c('Feature', 'Name'),
                        by.x=c('Feature', 'Name')) %>%
    mutate(p.groups = cut(P.value*ifelse(Coefficient < 0, -1, 1),
                          breaks = c(-0.05, -0.01, -0.001, 0, 0.001, 0.01, 0.05),
                          labels = c("beta < 0, p < 0.05",
                                     "beta < 0, p < 0.01",
                                     "beta < 0, p < 0.001",
                                     "beta > 0, p < 0.001",
                                     "beta > 0, p < 0.01",
                                     "beta > 0, p < 0.05")))

mycolors <- c("beta < 0, p < 0.05" = "#B2182B",
              "beta < 0, p < 0.01" = "#EF8A62",
              "beta < 0, p < 0.001" = "#FDDBC7",
              "beta > 0, p < 0.001" = "#D1E5F0",
              "beta > 0, p < 0.01" = "#67A9CF",
              "beta > 0, p < 0.05" = "#2166AC",
              "NA" = "gray50")

maaslinplot <- ggplot(maas.tables.df, aes(x = Feature, y = reorder(Name, my.plot.order(Name)), fill = beta)) +
    geom_tile(aes(fill = 1)) +
    geom_tile(colour="white", size=0.25) +
    scale_fill_distiller(palette = "RdBu", name = "Effect size",
                         limits=c(-0.003, 0.003), breaks = c(-0.003, 0, 0.003), na.value="grey30") +
    coord_fixed() +
    xlab("") +
    ylab("") +
    theme_classic(10) +  
    theme(axis.text.x = element_text(angle = 90, hjust = 1.0, vjust=0.5, size=10),
          axis.text.y = element_text(size=10))

ggsave(file = "cache/maaslin.png", plot = maaslinplot, height = 3, width = 4, unit = "in", dpi = 300)
```

Max value of abs beta is `r maas.results.df.signf %>% pull(Coefficient) %>% abs %>% max`.

```{r maaslin table}
maas.results.df.signf %>%
    select(Feature, Value, Coefficient, Qval) %>%
    arrange(Qval) %>%
    kable
```

```{r multi model, eval = FALSE, include = FALSE}
abund.genus.systm <- pseq.genus %>%
    abundances(., transform = "compositional") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id")

df.genus.systm <- meta(pseq.genus) %>%
    tibble::rownames_to_column("id") %>%
    full_join(., abund.genus.systm, by = "id") %>%
    mutate(Anaerotruncus = scale(asin(sqrt(g_Anaerotruncus))),
           Sutterella = scale(asin(sqrt(g_Sutterella))))

fo.genus.systm <- sprintf("oSYSTM ~ Anaerotruncus + Sutterella + Anaerotruncus:Sutterella + %s",
                          paste0(var.CL, collapse = " + "))

lm(as.formula(fo.genus.systm), data = df.genus.systm) %>% summary

df.genus.systm %>%
    mutate(diff = -0.75418 * Anaerotruncus + 0.89265 * Sutterella) %>%
    pull(diff) %>%
    summary
```

```{r scatter plots, eval = FALSE, include = FALSE}
scatter.signf.bacteria <- maas.results.df.signf %>% pull(Feature) %>% unique

scatter.meta <- meta(pseq.genus) %>%
    tibble::rownames_to_column("id") %>%
      mutate(MAP = oMAP,
           PULSEPRESSURE = oPULSEPRESSURE,
           SYSTM = oSYSTM,
           DIASM = oDIASM)

scatter.abund <- abundances(pseq.genus, transform = "clr") %>%
    t %>%
    as.data.frame %>%
    tibble::rownames_to_column("id") %>%
    select(id, scatter.signf.bacteria) %>%
    mutate_at(.vars = vars(scatter.signf.bacteria), .funs = funs(myscale(.)))

scatter.full <- full_join(scatter.meta, scatter.abund, by = "id")

scatter.combinations <- expand.grid(cc("SYSTM", "DIASM", "PULSEPRESSURE", "MAP"),
                                    c2l(scatter.signf.bacteria)) %>%
    setNames(c("rr", "abund"))

scatter.plot <- function(df,
                         maas.results,
                         rr,
                         abund,
                         covariates = c("BL_AGE", "SEX", "BMI", "CURR_SMOKE", "Q57X", "PREVAL_DIAB",
                                        "BL_USE_RX_C03", "BL_USE_RX_C07", "BL_USE_RX_C08",
                                        "BL_USE_RX_C09"),
                         prediction = TRUE,
                         plots = TRUE) {
    fo <- sprintf("%s ~ %s + %s", rr, abund, paste0(covariates, collapse="+"))
    model <- lm(as.formula(fo), data = df)

    if (prediction == FALSE) return(model %>% summary)

    df.predict <- data.frame(BPV = seq(-5,5,0.1)) %>%
        mutate(SEX = as.factor(0),
               BL_AGE = df %>% pull(BL_AGE) %>% mean,
               BMI = df %>% pull(BMI) %>% mean,
               CURR_SMOKE = as.factor(0),
               Q57X = as.factor(1),
               PREVAL_DIAB = as.factor(0),
               BL_USE_RX_C03 = as.factor(0),
               BL_USE_RX_C07 = as.factor(0),
               BL_USE_RX_C08 = as.factor(0),
               BL_USE_RX_C09 = as.factor(0),
               !!abund := seq(-5,5,0.1)
               )
    prediction <- cbind(df.predict, predict(model, df.predict, interval = 'confidence'))

    if (plots == FALSE) return(prediction)

    feature <- replace.brackets(abund)

    qval.maaslin <- maas.results %>%
        dplyr::filter(Feature == feature, Value == rr) %>%
        nrow

    if (qval.maaslin == 0) { return (ggplot() + ggplot2::theme_minimal()) }
    pval <- pub.p(summary(model)$coef[2,4])
    rsqr <- signif(summary(model)$adj.r.squared, 2)
    slope <- signif(model$coef[[2]], 2)

    ggplot(prediction, aes_string(x = abund, y = "fit")) +
        geom_line() +
        geom_point(size=0.1, shape=1, aes_string(x = abund, y = rr, color = "SEX"),
                   data = df) +
        labs(title = sprintf("Adj R2 = %s, Slope = %s, p = %s", rsqr, slope, pval)) +
        xlab(feature) +
        ylab(rr) +
        ylim(0,200) +
        ggplot2::theme_minimal() +
        theme(plot.title = element_text(size=8)) +
        theme(legend.position="none")
}

scatterplots <- lapply(data.frame(t(scatter.combinations)),
                       function(x)
                           scatter.plot(scatter.full, maas.results.df.signf, x$rr, x$abund))

scatterplot <- do.call("grid.arrange", c(scatterplots, ncol=4))
ggsave(file = "cache/scatter.png", plot = scatterplot, height = 10, width = 10, dpi = 300)
```

```{r maaslin means, eval = FALSE}
pseq.genus.core.table <- pseq.genus.core %>%
    prune_taxa(c("g_Adlercreutzia", "g_Anaerotruncus", "g_Sutterella"), .) %>%
    transform(transform = "compositional") %>%
    psmelt %>%
    mutate(map_groups = relevel(factor(dplyr::ntile(MAP, 3)), ref='1'),
           systm_groups = relevel(factor(dplyr::ntile(SYSTM, 3)), ref='1'),
           Abundance = myscale(Abundance))

table_bp_abund <- function(pseq, group) {
    pseq %>%
        group_by(.dots = c(group, "OTU")) %>%
        summarize(mean = mean(Abundance), sd = sd(Abundance)) %>%
        mutate(beta_sd = sprintf("%.2f ± %.2f", mean, sd)) %>%
        select(OTU, beta_sd) %>%
        spread(OTU, beta_sd) %>%
        ungroup %>%
        dplyr::rename("grp" = group) %>%
        mutate("grp" = paste0(group, "_", grp))
}

rbind(table_bp_abund(pseq.genus.core.table, "map_groups"),
             table_bp_abund(pseq.genus.core.table, "systm_groups")) %>%
    tibble::column_to_rownames("grp") %>%
    t %>%
    as.data.frame %>%
    kable

```{r Write figure 3, include=FALSE}
fig3head <- "Associations between genus level abundances and blood pressure variables."
fig3foot <- "Only bacteria with significant associations are shown. Blood pressure variables are normalized and abundances are arcsin-square root transformed. Dark gray color represents insignificant associations."
writeimage(doc, 3, "cache/maaslin.png", fig3head, fig3foot)
```

>>>>

@@

```{r deseq continuous}
dset.deseq.signf <- deseqresults(dds, names.dset)
```

```{r deseq heatmap}
g.deseq <- deseqhaetmap(dset.deseq.signf)
ggsave(file = "cache/deseq.png", plot = g.deseq, height = 3, width = 12, unit = "in", dpi = 300)
```

```{r deseq abundances for boxplot}
dset.deseq.boxplot <- pseq.genus %>%
    microbiome::transform(., transform = "compositional") %>%
    prune_taxa(unique(pull(dset.deseq.signf, Feature)), .) %>%
    psmelt %>%
    mutate(Abundance = sqrt(Abundance))
```

```{r deseq boxplot}
g.boxplot <- dset.deseq.boxplot %>%
    group_by(OTU) %>%
    mutate(outlier = abs(Abundance) > median(Abundance) + IQR(Abundance) * 1.5) %>%
    ggplot(aes(x = HYPERTENSION, y = Abundance, color = OTU)) +
    geom_boxplot(lwd=0.1, outlier.shape = NA) +
    geom_point(data = function(x) dplyr::filter_(x, ~ outlier),
               position = position_jitterdodge(jitter.width = 0.02),
               size = 0.1,
               alpha = 0.4, shape = ".") +
    scale_y_continuous(limits = c(0, 0.2), expand = c(0, 0)) +
    scale_x_discrete(labels=c("0" = "Normotensive", "1" = "Hypertensive")) +
    ylab("Abundance (Hellinger)") +
    xlab("") +
    theme_classic() +
    guides(shape = guide_legend(override.aes = list(size = 0.1)),
           color = guide_legend(override.aes = list(size = 0.1))) +
    theme(legend.title = element_text(size = 4), 
          legend.text  = element_text(size = 4),
          legend.key.size = unit(1, "lines"),
          legend.position="bottom")

ggsave(file = "cache/dds-test.png", plot=g.boxplot, height=4, width=6, units = "in", dpi = 300)
```

```{r Write figure 3, include=FALSE}
fig3head <- "Associations between genus level abundances and blood pressure variables."
fig3foot <- "Only bacteria with significant associations are shown. Blood pressure variables are normalized and abundances are arcsin-square root transformed. Dark gray color represents insignificant associations."
writeimage(doc, 3, "cache/deseq.png", fig3head, fig3foot)
```
